Algebraic Groups\vspace{-1ex}

### 2019-12-02 Algebraic Geometry/Algebraic-Geometry.tex
update

>>>>

\section{December 2nd, 2019}
Today (and through the rest of the week) we are going to talk about sheaves! Woot.

We're going to have a long time in between our last class this week and our first class in January. Here's a tidbit to tide you over:
\begin{prob}
	Homework for the break: Do all the problems in section 1 of chapter II of Hartshorne.
\end{prob}
\subsection{Sheaves}
There is a mantra! \textit{Sheaves are a geometric manifestation of the flow of information from local to global.}
\begin{ex}
	Let $X\in\SpcFun$. Then $\O$ is the sheaf of regular functions on $X$.
\end{ex}
\begin{rmk}
	Recall that we had that $\O(U)$ was defined for every open $U$ and that regulartity was preserved by 
	restrictions and finally that regularity is local (it glues). These give us the ideas we will use to define a sheaf.
\end{rmk}
I love this little gluing diagram: given an open covering $U_i$ of $U$,
\begin{center}
	\begin{tikzcd}
		\O(U)\ar[r] & \prod_i \O(U_i)\ar[r,shift left]\ar[r,shift right] & \prod_{i,j}\O(U_i\cap U_j)
	\end{tikzcd}
\end{center}
where the top map sends
\[(\varphi_i)_i\mapsto (\varphi_i|_{U_i\cap U_j})_{(i,j)}\]
and the bottom does 
\[(\varphi_i)_i\mapsto (\varphi_j|_{U_i\cap U_j})_{(i,j)}.\]
That is, $\O(U)$ is the equalizer of these maps.

We took some time to talk about limits as (right) adjoints to the constant diagram functor. Hadn't seen that before.
We concluded with a bunch of cateory theory, which was really fun. Ind and Pro categories.
\subsection{More examples}
Let either
\begin{itemize}
	\item $X$ and $Y$ be spaces
	\item $h_Y:U\mapsto h_Y(U)$ is a sheaf on $X$ for all open $U$ in $X$.
\end{itemize}

### 2019-12-02 Algebraic Groups/Algebraic-Groups.tex
update

	So there exists a projection map $\pi:A\to k[\varepsilon]/\varepsilon^2\cong k\oplus k\varepsilon$ that sends 
	\[x\mapsto \langle x,c\rangle+\varepsilon\langle x,d\rangle\]
	for some $c,d\in C$.
	Then a subclaim is to prove that
	\begin{itemize}
		\item $\Delta(c)=c\otimes c$
		\item $\Delta(d)=c\otimes d+d\otimes c$
		\item $\varepsilon(c)=1$
	\end{itemize}
	Then by also showing that $c$ is invertible, one can compute 
	\[\Delta(dc^{-1})=1\otimes c^{-1}d+c^{-1}d\otimes 1\]
	which then gives us a (nontrivial!) map $G\to \Ga$, which contradicts the assumption.
\end{prf}
\begin{ex}[Non co\'etale example]
	Consider $\Ga=\Spec k[x]$ with $\delta(x)=x\otimes 1+1\otimes x$. Then consider th subcoalgebra spanned by $\langle 1,\dots,x^{n-1}\rangle$
	and the dual algebra $A=C^\vee$ spanned by $e_i$. One can compute that 
	\[e_i\cdot e_j=\binom{i+j}{i}e_{i+j}\]
	if $i+j\le n-1$ and it is zero otherwise. But then in particular $e_{n-1}^2=0$, so $A$ isn't even reduced!

@@

>>>>

	So there exists a projection map $\pi:A\to k[\varepsilon]/\varepsilon^2\cong k\oplus k\varepsilon$ that sends 
	\[x\mapsto \langle x,c\rangle+\varepsilon\langle x,d\rangle\]
	for some $c,d\in C$.
	Then a subclaim is to prove that
	\begin{itemize}
		\item $\Delta(c)=c\otimes c$
		\item $\Delta(d)=c\otimes d+d\otimes c$
		\item $\varepsilon(c)=1$
	\end{itemize}
	Then by also showing that $c$ is invertible, one can compute 
	\[\Delta(dc^{-1})=1\otimes c^{-1}d+c^{-1}d\otimes 1\]
	which then gives us a (nontrivial!) map $G\to \Ga$, which contradicts the assumption.
\end{prf}
\begin{ex}[Non co\'etale example]
	Consider $\Ga=\Spec k[x]$ with $\Delta(x)=x\otimes 1+1\otimes x$. Then consider th subcoalgebra spanned by $\langle 1,\dots,x^{n-1}\rangle$
	and the dual algebra $A=C^\vee$ spanned by $e_i$. One can compute that 
	\[e_i\cdot e_j=\binom{i+j}{i}e_{i+j}\]
	if $i+j\le n-1$ and it is zero otherwise. But then in particular $e_{n-1}^2=0$, so $A$ isn't even reduced!

@@

\section{December 2nd, 2019}
We are going to continue our discussion on Jordan decomposition by visiting the idea of 
\subsection{Unipotent group schemes}
Recall that if $G$ is a linear algebraic group over $k$ and $g\in G(k)$, then 
\begin{itemize}
	\item $g$ is semisimple if $\rho(g)$ is for all representations $\rho:G\to \GL(V)$ ($\rho(g)\otimes \bar k$ is diagonalizable)
	\item $g$ is unipotent if $\rho(g)$ is for all representations $\rho$ ($\rho(g)-1$ is nilpotent).
\end{itemize}

We also said that, if $k$ is perfect, $g=g_sg_u=g_ug_s$ and this is a unique decomposition (the Jordan decomposition).

When we are talking about an arbitrary group scheme, we no longer use ``semisimple'' but rather ``multiplicative type'': 
If $G\subseteq \Gm^n\subseteq\GL_n$, then we say that $G$ is \textbf{diagonaliable}. If $G_{\bar k}$ is diagonalizable, then 
we say $G$ is \textbf{of multiplicative type.}

Recall that when $G$ is smooth, we saw that $G$ was of multiplicative type if and only if $G$ is commutative and $G(\bar k)$ consists entirely of semisimple objects.

Some questions we want to answer: Is there a characterization of groups containing unipotent elements? Perhaps more interestingly:
Do we have structure results for factoring algebraic groups analogous to the Jordan decomposition?

\begin{thm}[Kolchin fixed point theorem]
	Let $V$ be a finite dimensional vector space over $k$ and let $G\subseteq \GL(V)$ be an abstract subgroup containing unipotent elements.
	Then there exists a nonzero $v\in V$ fixed by all $g\in G$.
\end{thm}
\begin{prf}
	Begin by setting 
	\[V^G=\{v\in V|\forall g\in G, gv=v\}.\]
	It is easy to see that if $k'/k$ is a field extension, then 
	\[(V\otimes_k k')^G=V^G\otimes_k k'\]
	thus we can assume that $k=\bar k$ (since we want to show that $V^G$ is nonempty).

	We can also assume that $V$ is a simple $G$-module, since it suffices to find a fixed point in a submodule.
	This reduces the proof to showing that $V^G=V$.

	There are two different proofs of this: first, we want to show that for all $g\in G$, that $g-I_V=0$. Suppose there were 
	an element $g'$ not equal to the identity. We know that for all $g\in G$, (since $g$ is unipotent):
	\[\tr (g)=\dim V\quad\Rightarrow\quad \tr(gg')=\dim V\quad\Rightarrow\quad \tr(g(g'-I_V))=0\]

	Let $E=\{f:V\to V|\tr(gf)=0,\forall g\in G\}$, which contains $g'-I_V\ne 0$. $E$ is $G$-invariant under composition.
	Then let $g'-I_V\in X\subseteq E$ where $X$ is a simple $G$-module containing this element. THen for any $v\in V$, define 
	\[\varphi_v:X\to V\quad\text{via}\quad f\mapsto f(v)\]
	because $g'\ne I_V$, there exists a $v\in V$ such that $g'v\ne v$. Thus for this specific $v$, $\varphi_v$ is an isomorphism!

	In particular, this means there exists an $f\in X$ such that $f(v)=v$. THen take 
	\[A=\{L:V\to V| L\text{ linear, commuting with $G$}\}\]
	which is a division ring over $k=\bar k$, which forces $A=k$.

	But then for all $w\in V$, we get the composition 
	\begin{center}
		\begin{tikzcd}
			V\ar[r,"\phi_v^{-1}"]\ar[rr,bend right] & X\ar[r,"\phi_w"] & V
		\end{tikzcd}
	\end{center}
	and then we are supposed to show that our choice of $f$ and $v$ above send all other $w\in V$ to the span of $v$,
	so $f$ has trace one. Something is fishy in this proof, but it's in Waterhouse.
\end{prf}
\begin{prf}[Less tricky, more fancy]
	By Wedderburn, if $A$ is a ring and $M$ is a faithful left $A$-module, and if $B=\End_A(M)$, then if $M$ is simple and finitely generated over $B$, 
	then 
	\[\End_B(M)=A.\]
	The more traditional setting here is when $A=k$, $M=k^n$, and $B=M_n(k)$. Then $\End_{M_n(k)}(k^n)=k.$

	Now let $A\subseteq \End_k(V)$ generated by $G$. Let $M=V$ (which is assumed to be simple as above). Then by letting $B=\End_A(V)=k$, so by the above 
	result, $\End_k(V)=A$ so $A$ is simple! Then let $I=\langle g-I_V:g\in G\rangle$, a two-sided ideal of $A$, so is either zero (in which case we are done) or is all of $A$.

\end{prf}
